humanCdGF_predict_cfu

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
refine feature selection

seed <- 1
n_trees <- 1001

@@

##### feature reduction #####
### boruta
library(Boruta)
set.seed(seed)
rf_features <- Boruta(log_cfu_med ~ ., data = cfu_shared_colonized.df, maxRuns = 2000)
boruta_otus <- names(rf_features$finalDecision)[rf_features$finalDecision == 'Confirmed']
set.seed(seed)
rf_boruta <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', boruta_otus)],
                           importance = TRUE, ntree = n_trees)
rf_bor_import <- sort(importance(rf_boruta)[,1], decreasing = T)
detach('package:Boruta', unload = T)
detach('package:ranger', unload = T)
detach('package:randomForest', unload = T)
library(randomForest)
### above mean features, reiteratively reduce to 2 otus 
###### used mean because median results in including many low values in early OTU selection steps
rf_rsq <- c()
otus_used <- c()
rf_otus <- colnames(cfu_shared_colonized.df)[-1]
for(i in 1:100){
  if(length(rf_otus) > 1){
  set.seed(seed)
  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
                     importance = TRUE, ntree = n_trees)
  rf_import_otus <- importance(rf, type = 1)
  otus_used[[i]] <- rf_otus
  rf_otus <- rownames(rf_import_otus)[rf_import_otus >= mean(rf_import_otus)]
  rf_rsq[i] <- tail(rf$rsq, 1)
  } else {
    set.seed(seed)
    rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
                       importance = TRUE, ntree = n_trees)
    otus_used[[i]] <- rf_otus
    rf_rsq[i] <- tail(rf$rsq, 1)
    stop('Only one OTU remains')
  }
}
#### select top otus from reiteration through RF
optimum_otus <- otus_used[[which(rf_rsq == max(rf_rsq))]]
reduced_otus <- otus_used[[6]]

### test models with top subset of otus
top_important_OTU <- reduced_otus
top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
  select(log_cfu_med, one_of(top_important_OTU)))

@@

#One per cage
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, Otu00250, Otu00005))

om_rsq_rf <- c() # rsq of each rf model
om_rsq <- c() # cumulative rsq of all models
om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
one_df <- c()
om_pred <- c()
for(i in 1:500){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_rsq_rf[i] <- tail(rf_om$rsq, 1)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test)))
    om_rsq[i] <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_import_otus <- select(om_import_otus, -IncNodePurity)
}
test <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(test) <- seq(ncol(test))
### rank
test_rank <- apply(test, 2, function(x) rank(x))
test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
### mean
test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
#### median
test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)

#####
test <- c() # test if any model has same train set
  for(i in 2:21){
  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
}
any(test) # if true, then trained on the same set of mice multiple times
#####

@@

med_otu_importance_om <- data.frame(otu = om_import_otus$otu,
                                 importance = apply(om_import_otus[, -1], 1, median))
med_otu_importance_om <- med_otu_importance_om[order(med_otu_importance_om$importance, decreasing = T),]

@@

ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + geom_point() + 
  geom_abline(slope = 1) + 
  xlim(c(4,9)) + ylim(c(4,9)) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
  theme_bw() + annotate('text', x=1.05*(min(LOcO_probs$observed)), 
                        y=0.95*(max(LOcO_probs$predicted)), # parse = T,
                        label=paste0('OOB ', 'R^2 == ', round(tail(rf_n_features$rsq, 1), 2),
                                     '\nLOcO ','R^2 == ', round(LOcO_rsq, 2),
                                     '\n1 mouse', 'R^2 == ', round(tail(om_rsq, 1), 2))) + 
  theme(legend.position = 'none')

>>>>

seed <- 1
n_trees <- 1001
n_otus <- 12
iters <- 100
```

@@

##### feature reduction #####
###### select top n otus reiteratively and select otus that appear > 50% of iterations
rf_n_otus <-lapply(1:iters, function(i){
  set.seed(i)
  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
                     importance = TRUE, ntree = n_trees)
  names(sort(importance(rf)[,1], decreasing = T))[1:n_otus]
})
n_otu_freq <- table(unlist(rf_n_otus))
top_important_OTU <- names(n_otu_freq)[n_otu_freq >= iters/2]

top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
  select(log_cfu_med, one_of(top_important_OTU)))

@@

#One per cage
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))

om_rsq_rf <- c() # rsq of each rf model
om_rsq <- c() # cumulative rsq of all models
om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
om_pred <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_rsq_rf[i] <- tail(rf_om$rsq, 1)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test)))
    om_rsq[i] <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_import_otus <- select(om_import_otus, -IncNodePurity)
}
test <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(test) <- seq(ncol(test))
### rank
test_rank <- apply(test, 2, function(x) rank(x))
test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
### mean
#test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
#test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
#test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
#### median
#test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
#test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
#test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)

#####
#test <- c() # test if any model has same train set
#  for(i in 2:21){
#  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
#}
#any(test) # if true, then trained on the same set of mice multiple times
#####

@@

med_otu_importance_om <- data.frame(otu = names(test_import_rank),
                                 importance = test_import_rank)

@@

ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + geom_point() + 
  # geom_abline(slope = 1) + 
  xlim(c(4,9)) + ylim(c(4,9)) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
  theme_bw() + annotate('text', x = 4.5, y = 8.5, # parse = T,
                        label=paste0('OOB ', 'R^2 == ', round(tail(rf_n_features$rsq, 1), 2),
                                     '\nLOcO ','R^2 == ', round(LOcO_rsq, 2),
                                     '\n1 mouse', 'R^2 == ', round(tail(om_rsq, 1), 2))) + 
  theme(legend.position = 'none')
